Open Clusters Characterization in Gaia DR2 Using ML Algorithms

### 2020-12-07 thesis/thesis.tex
thesis: Add Cluster Characterization with VO Tools chapter

% TODO: Fix table if figures changes
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{l|c|c|c|c}
      \textbf{Source / method} & \textbf{$\mu_{\alpha}$ $(mas/yr^{-1})$} & \textbf{$\mu_{\delta}$ $(mas/yr^{-1})$} & \textbf{$\varpi$ $(mas)$} & \emph{\# stars} \\
      \hline
      Simbad.u-strasbg.fr\tablefootnote{Results have been taken from \protect\citeA[p.~25. Table A.3. Pleiades]{babusiaux2018gaia}} &
        19.997 $\pm$ 0.127 & -45.548 $\pm$ 0.101 & 7.364 $\pm$ 0.005 & 1326 \\
      Clusterix+TOPCAT & 19.98 $\pm$ 1.25 & -45.47 $\pm$ 1.48 & 7.33 $\pm$ 0.21 & 168 \\
      KMeans & 20.25 $\pm$ 35.16 & -38.01 $\pm$ 40.16 & 7.23 $\pm$ 2.35 & 1378 \\
      DEC & 23.67 $\pm$ 38.33 & -46.23 $\pm$ 44.54 & 8.04 $\pm$ 2.59 & 878 \\
      DEC (filtered) & 19.50 $\pm$ 8.59 & -44.23 $\pm$ 8.17 & 7.42 $\pm$ 0.1 & 438 \\
    \end{tabular}
    \caption{Melotte 22. The parameters shown are proper motion in right ascension and declination, parallax and number of stars with their respective deviations.}
    \label{tab:results_melotte_22}
  \end{center}
\end{table}

@@

We have seen in the previous chapter that the DEC model is able to refine the clustering made by K-Means. This allows us to get
a more accurate characterization of the cluster.

In order to validate our model and be able to say how good or bad it is we need a different method to characterize the cluster
and compare one against each other.

For that reason we have used tools from the Virtual Observatory (VO) such as Clusterix and TOPCAT to characterize the studied clusters.

@@

>>>>

% TODO: Fix table if figures changes
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{l|c|c|c|c}
      \textbf{Source / method} & \textbf{$\mu_{\alpha}$ $(mas/yr^{-1})$} & \textbf{$\mu_{\delta}$ $(mas/yr^{-1})$} & \textbf{$\varpi$ $(mas)$} & \emph{\# stars} \\
      \hline
      \textbf{Simbad.u-strasbg.fr}\tablefootnote{Results have been taken from \protect\citeA[p.~25. Table A.3. Pleiades]{babusiaux2018gaia}} &
        19.997 $\pm$ 0.127 & -45.548 $\pm$ 0.101 & 7.364 $\pm$ 0.005 & 1326 \\
      Clusterix+TOPCAT & 19.98 $\pm$ 1.25 & -45.47 $\pm$ 1.48 & 7.33 $\pm$ 0.21 & 634 \\
      KMeans & 20.25 $\pm$ 35.16 & -38.01 $\pm$ 40.16 & 7.23 $\pm$ 2.35 & 1378 \\
      DEC & 23.67 $\pm$ 38.33 & -46.23 $\pm$ 44.54 & 8.04 $\pm$ 2.59 & 878 \\
      \textbf{DEC (filtered)} & 19.50 $\pm$ 8.59 & -44.23 $\pm$ 8.17 & 7.42 $\pm$ 0.1 & 438 \\
    \end{tabular}
    \caption{Melotte 22. The parameters shown are proper motion in right ascension and declination, parallax and number of stars with their respective deviations.}
    \label{tab:results_melotte_22}
  \end{center}
\end{table}

@@

We have seen in the previous chapter that the DEC model is capable of refining the clustering selection made by K-Means.
This allows us to obtain a more precise characterization of the cluster.
Now, in order to validate our model and be able to estimate how good or bad it is,
we need a different method to characterize the studied clusters and compare our results with the validation method.

For that reason, we have used an alternative method based on Virtual Observatory (VO) tools such as Clusterix and TOPCAT,
to characterize the studied clusters.

@@

The aim of this section is to highlight procedure differences between the our method and this new one,
as well as using the results given by this method to validate ours results.
This validation method uses a combination of VO tools: Clusterix 2.0 + TOPCAT.
We take Melotte 22 again as our study case to make easier taking comparisons between the method presented in Chapter \ref{chap:method}
and this alternative method.

The initial point, in order to avoid a subjective and manual selection for the cluster members, is taking advantage of Clusterix.
Figure \ref{fig:clusterix_selection} shows a screenshot of the Clusterix 2.0 selection panel.

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.9\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../figures/clusterix/clusterix_selection_panel_melotte_22.png}
  \end{subfigure}
  \caption{Clusterix 2.0 selection panel with Melotte 22 settings.
           The coordinates correspond to the center of Melotte 22 cluster,
           while the radius is the value registered for this cluster in the
           OpenClust catalogue multiplied by 1.5 (as explained in \ref{sec:data_mining})}
  \label{fig:clusterix_selection}
\end{figure}

This panel allows us to select a region by its coordinates or by searching a cluster by its name in the Gaia DR2 catalogue,
and specify the size of the region by introducing its radius.
We can also set limit values in magnitude (for both, minimum and maximum) to discard stars outside those limits.
Another setting is the flag to enable or disable the \emph{Q-Filter}.
This filter discards those stars with \emph{null} values in relevant fields.

Clusterix returns a position map for the downloaded stars from Gaia DR2 database using Aladin \cite{bonnarel2000aladin} as an atlas of the sky
(See Figure \ref{fig:aladin_melotte_22}).
At this point, we can choose sending the recovered data to TOPCAT or Aladin, or continue to the next step of this assistant.

We select the second choice and proceed with the proper motion frequency space analysis.

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.9\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../figures/clusterix/clusterix_aladin_selection_melotte_22.png}
  \end{subfigure}
  \caption{Aladin screenshot with equatorial coordinates for Melotte 22 stars referred to J2000.}
  \label{fig:aladin_melotte_22}
\end{figure}

In this step, we define three regions that will be used for the frequency analysis:

\begin{itemize}
  \item \textbf{Inner region} (\emph{c+f}), with stars that belong to the OC and field stars.
  \item \textbf{External region} or field region (\emph{f}), with stars that do not belong the open cluster.
  \item \textbf{Void region}. This region is excluded from the analysis since it is supposed to be a transition region with stars that could belong the OC.
\end{itemize}

It is necessary to notice the importance of this step, since the selection of these regions is critical to get a good result.
However, despite how crucial is this selection, there is no general rule to define these regions.
We only have two references: the radius of the downloaded region and the estimated size for the open cluster that sets the \emph{inner radius}.
The other two radius must be set by trial and error and, in general, they must be generous to be sure we are not taking polluted regions.
This is a tedious process that requires several tries before achieving good enough probability values.

Furthermore, it is necessary to know in advance some additional profile of the cluster, in addition to the estimated size.
For example, in the case of Melotte 22, the proper motion in declination direction of the membership stars is around $-45 mas \cdot year^{-1}$,
so it is necessary to open the $\mu$ range above the estimated values (in absolute value). If we do not do that we loose its detection,
since Clusterix sets this value to $30 mas \cdot year^{-1}$ by default. But even in this case, it is necessary to set a threshold for
magnitude values, $G \leq 14$, to help getting the right result.
Figure \ref{fig:clusterix_step2_melotte_22} shows the region selection panel on Clusterix 2.0.

This makes the selection process an art rather than a science, not to mention the required previous knowledge on certain profiles of the
objects to characterize. This puts into question the non-parametric character of the method.

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.9\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../figures/clusterix/clusterix_step2_melotte_22.png}
  \end{subfigure}
  \caption{Clusterix 2.0 Step 2/3: Region selection.}
  \label{fig:clusterix_step2_melotte_22}
\end{figure}

After having set all these parameters, Clusterix performs the proper motion frequency analysis and returns the probability field
associated to each star of the whole region (see Figure \ref{fig:clusterix_probability_melotte_22}).

This step can be repeated over and over again, varying radii and other parameters, until finally reach a valid probability that works.
The last step presents a table with candidates objects and allows to export this dataset to TOPCAT, Aladin and VOSA to perform a deeper
filtering and analysis based on the results given by Clusterix.

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.9\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../figures/clusterix/clusterix_probability_melotte_22.pdf}
  \end{subfigure}
  \caption{Kinematic probability associated to the membership of the open cluster.
           This probability is derived from the proper motion frequency analysis.}
  \label{fig:clusterix_probability_melotte_22}
\end{figure}

Once data is available in TOPCAT, we have to carefully analyze the imported data.
From there, we can perform a first stars selection by excluding those stars with probability value lower than 0.8.
This first result can be taken as a good staring point, but we still need to make some fine tunning.
As we can see in Figure \ref{fig:topcat_prob_selection_melotte_22}, the selection is too restrictive and we have lost too much stars.

Looking at Figure \ref{fig:topcat_prob_parallax_melotte_22}, we can assume that cluster members are around $7.3mas$ in the parallax histogram.
Also, Figure \ref{fig:topcat_prob_pm_melotte_22} reveals that cluster stars have low scattering rate in the proper motion configuration space.

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.9\textwidth}
    \centering
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/clusterix/topcat_prob_pm_melotte_22.png}
      \caption{Proper motion}
      \label{fig:topcat_prob_pm_melotte_22}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/clusterix/topcat_prob_parallax_melotte_22.png}
      \caption{Parallax}
      \label{fig:topcat_prob_parallax_melotte_22}
    \end{subfigure}
  \end{subfigure}
  \caption{Melotte 22 first selection using validation method. Parallax is $7.304mas$ with 0.797 of standard deviation.}
  \label{fig:topcat_prob_selection_melotte_22}
\end{figure}

Taken these observations into account, we can proceed to make a new selection based on parallax properties.
Our second selection uses the values obtained with the previous selection for parallax and its standard deviation and
selects those stars from the whole set which fall inside the limits $7.304mas \pm 0.797$ (see Figure \ref{fig:topcat_parallax_melotte_22}).

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.9\textwidth}
    \centering
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/clusterix/topcat_2nd_selection_pm_melotte_22.png}
      \caption{Proper motion}
      \label{fig:topcat_2nd_selection_pm_melotte_22}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/clusterix/topcat_2nd_selection_parallax_melotte_22.png}
      \caption{Parallax}
      \label{fig:topcat_2nd_selection_parallax_melotte_22}
    \end{subfigure}
  \end{subfigure}
  \caption{Melotte 22 second selection. The estimated number of cluster members is 709.}
  \label{fig:topcat_parallax_melotte_22}
\end{figure}

Looking at the H-R diagrams (Figure \ref{fig:clusterix_selection_hr_diagrams_melotte_22}),
we can compare both selections and see how we have extended the selection without getting too much noise around the isochrone curve.

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.9\textwidth}
    \centering
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/clusterix/topcat_1st_selection_hr_diagram_melotte_22.png}
      \caption{H-R diagram from first selection.}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/clusterix/topcat_2nd_selection_hr_diagram_melotte_22.png}
      \caption{H-R diagram from second selection.}
    \end{subfigure}
  \end{subfigure}
  \caption{Comparison between first and second selection H-R diagrams.}
  \label{fig:clusterix_selection_hr_diagrams_melotte_22}
\end{figure}

Finally, we filter the second selection by discarding those stars whose parallax relative error in absolute value is behind 0.05
(see Figure \ref{fig:clusterix_final_selection_melotte_22}).

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.9\textwidth}
    \centering
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/clusterix/topcat_3rd_selection_parallax_melotte_22.png}
      \caption{Parallax}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/clusterix/topcat_3rd_selection_hr_diagram_melotte_22.png}
      \caption{H-R diagram}
    \end{subfigure}
  \end{subfigure}
  \caption{Comparison between first and second selection H-R diagrams.}
  \label{fig:clusterix_final_selection_melotte_22}
\end{figure}

For the sake of completeness, Table \ref{tab:results_melotte_22} is shown again, in order to compare the results obtained with the validation method and our method.

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{l|c|c|c|c}
      \textbf{Source / method} & \textbf{$\mu_{\alpha}$ $(mas/yr^{-1})$} & \textbf{$\mu_{\delta}$ $(mas/yr^{-1})$} & \textbf{$\varpi$ $(mas)$} & \emph{\# stars} \\
      \hline
      \textbf{Simbad.u-strasbg.fr} & 19.997 $\pm$ 0.127 & -45.548 $\pm$ 0.101 & 7.364 $\pm$ 0.005 & 1326 \\
      \textbf{Clusterix+TOPCAT} & 19.98 $\pm$ 1.25 & -45.47 $\pm$ 1.48 & 7.33 $\pm$ 0.21 & 634 \\
      KMeans & 20.25 $\pm$ 35.16 & -38.01 $\pm$ 40.16 & 7.23 $\pm$ 2.35 & 1378 \\
      DEC & 23.67 $\pm$ 38.33 & -46.23 $\pm$ 44.54 & 8.04 $\pm$ 2.59 & 878 \\
      DEC (filtered) & 19.50 $\pm$ 8.59 & -44.23 $\pm$ 8.17 & 7.42 $\pm$ 0.1 & 438 \\
    \end{tabular}
  \end{center}
\end{table}
